Simulating physiological flexibility in the acute glucocorticoid response to stressors reveals limitations of current empirical approaches

### 2022-07-01 3_r_scripts/Revision.Rmd
minor changes

The recognition that characterizing the functional shape of an acute stress response is challenging goes back to the earliest studies conducted in wild animals. Early studies often employed various control groups and sampled individual animals at a variety of time points over a long period in order to describe the full response curve for a particular group [e.g., a species or a breeding stage, @wingfield1992]. These validations were considered essential to characterize key parameters of the acute response for each group being studied (i.e., baseline, rate of increase, maximum level, time of peak, and area under the curve; John Wingfield, *personal communication*). The challenge of estimating these parameters becomes much more difficult when trying to describe the response for an individual animal rather than for a group, because glucocorticoids can often only be measured at two or three time points and only a small number of times per animal [e.g., @vitousek2018]. Because these studies require an estimate for each individual, the solutions used by older studies that added additional animals to allow for sampling at more time points are not available.

INSERT HISTORY AND CITATIONS FROM COCKREM

For individual based studies, the most common approach to this problem is to standardize measurements as much as possible by measuring animals at the same time of the day during the same context, and by taking blood samples at standard times (often <3 and 30 minutes after capture) to characterize baseline and stress-induced glucocorticoids. This standardization allows for comparison between individuals, but in some cases it may also completely obscure the ability to detect variation in certain characteristics of the acute response curve. For example, if the speed (rate of initial increase or time required to reach maximum) and scope (maximum value) of the acute response vary independently, samples taken at only two time points cannot accurately capture variation in either parameter [@taff2022]. Indeed, several discussions in recent years about methods such as the '3 minute rule' and the relative merits of 'area under the curve' versus time point measures of glucocorticoids are fundamentally related to a recognition of the importance of understanding variation in the functional shape of stress responses and whether different components of that shape covary within individuals [e.g., @cockrem2002; @small2017]. While a great deal of empirical work has focused on characterizing the rate of initial increase [CITE], there has been relatively little work on understanding individual differences in the time required to reach maximal levels, and this attribute of speed is harder to estimate [@taff2022].

@@

There is also a growing number of mathematical models of the  (Luttbeg, Taborsky)

@@

The goal of this simulation is to provide a flexible tool that can produce realistic datasets of physiological flexibility for a variety of different systems and scenarios. As such, there are many possible applications and here I briefly highlight a few possibilities. First, I demonstrate that the simulation can produce datasets that are qualitatively similar to empirical data on acute glucocorticoid responses. Next, I explore three specific scenarios with the simulation that highlight challenges to empirical progress in this field. First, I ask how well maximum glucocorticoid measures can be estimated with a single time point measure in populations that differ systematically in features of the response. Second, I explore how different patterns of covariation between maximum glucocorticoids and one measure of speed (the time to reach maximum) influence the ability to reliably measure either component of the response. Third, I ask how different amounts of between- and within-individual variation in maximum glucocorticoid levels impact the ability to detect known relationships between glucocorticoids and fitness. These questions are all important for understanding physiologically flexibility and are difficult to fully address empirically, but this is by no means an exhaustive list of the questions or possible parameter permutations that could be explored with simulation. Finally, I demonstrate how the simulation tools can be used to evaluate the performance of different experimental designs given a set of logistical limitations (e.g., number of samples that can be collected) and a proposed hypothesis) and discuss how this procedure could be used as a planning tool to increase the power and reproducibility of future empirical work.

>>>>

The recognition that characterizing the functional shape of an acute stress response is challenging goes back to the earliest studies conducted in wild animals. Early studies often employed various control groups and sampled individual animals at a variety of time points over a long period in order to describe the full response curve for a particular group [e.g., a species or a breeding stage, @wingfield1992]. These validations were considered essential to characterize key parameters of the acute response for each group being studied (i.e., baseline, rate of increase, maximum level, time of peak, and area under the curve; John Wingfield, *personal communication*). Indeed, there is a long and rich history of empirical work characterizing differences in each aspect of the acute glucocorticoid response [e.g., @cockrem2002; @wingfield1992; @breuner1999; @love2003] and empirical data has contributed to a variety of conceptual models of acute glucocorticoid regulation [e.g., @wingfield1998; @romero2009]. More recently, mathematical models have been used to generate predictions about how flexibility and variation in different parts of the response system might respond to particular conditions [e.g., @luttbeg2021; @grindstaff2022; @taborsky2021]. However, the challenge of estimating these parameters and of decomposing within- and between-individual variance becomes much more difficult when trying to fully describe the response for an individual animal rather than for a group, because glucocorticoids can often only be measured at two or three time points and only a small number of times per animal [e.g., @vitousek2018]. Because these studies require an estimate for each individual, the solutions used by older studies that added additional animals to allow for sampling at more time points are not available.

For individual based studies, the most common approach to this problem is to standardize measurements as much as possible by measuring animals at the same time of the day during the same context, and by taking blood samples at standard times (often <3 and 30 minutes after capture) to characterize baseline and stress-induced glucocorticoids. This standardization allows for comparison between individuals, but in some cases it may also completely obscure the ability to detect variation in certain characteristics of the acute response curve. For example, if the speed (rate of initial increase or time required to reach maximum) and scope (maximum value) of the acute response vary independently, samples taken at only two time points cannot accurately capture variation in either parameter [@taff2022]. Indeed, several discussions in recent years about methods such as the '3 minute rule' and the relative merits of 'area under the curve' versus time point measures of glucocorticoids are fundamentally related to a recognition of the importance of understanding variation in the functional shape of stress responses and whether different components of that shape covary within individuals [e.g., @cockrem2002; @small2017]. While a great deal of empirical work has focused on characterizing the rate of initial increase [@cockrem2005; @cockrem2006], there has been relatively little work on understanding individual differences in the time required to reach maximal levels, and this attribute of speed is harder to estimate [@taff2022].

@@

@@

The goal of this simulation is to provide a flexible tool that can produce realistic datasets of physiological flexibility for a variety of different systems and scenarios. As such, there are many possible applications and here I briefly highlight a few possibilities. Initially, I demonstrate that the simulation can produce datasets that are qualitatively similar to empirical data on acute glucocorticoid responses. Next, I explore three specific scenarios with the simulation that highlight challenges to empirical progress in this field. First, I ask how well maximum glucocorticoid measures can be estimated with a single time point measure in populations that differ systematically in features of the response. Second, I explore how different patterns of covariation between maximum glucocorticoids and one measure of speed (the time to reach maximum) influence the ability to reliably measure either component of the response. Third, I ask how different amounts of between- and within-individual variation in maximum glucocorticoid levels impact the ability to detect known relationships between glucocorticoids and fitness. These questions are all important for understanding physiologically flexibility and are difficult to fully address empirically, but this is by no means an exhaustive list of the questions or possible parameter permutations that could be explored with simulation. Finally, I demonstrate how the simulation tools can be used to evaluate the performance of different experimental designs given a set of logistical limitations (e.g., number of samples that can be collected) and discuss how this procedure could be used as a planning tool to increase the power and reproducibility of future empirical work.
